---
title: "Clean R code cheatsheet"
subtitle: Clean code in R workshop, eRum 2018
author: 
- "Ildi Czeller and Jeno Pal"
date: "May 2018"
output: 
  html_document:
    theme: united
    toc: true
---

<style type="text/css">

body{ /* Normal  */
      font-size: 16px;
  }
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 38px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 14px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```


_"One difference between a smart programmer and a professional programmer is that the professional understands that __clarity is king__. Professionals use their powers for good and write code that others can understand."_

# 1. Use meaningful names

## Use names that you can pronounce

```{r, class.source="bg-danger"}
rows_w_miss_val <- df[!complete.cases(df), ]
```

```{r, class.source="bg-success"}
rows_with_missing_values <- df[!complete.cases(df), ]
```

## Use names that are easy to distinguish

Avoid confusion stemming from using both single and plural of the same name

```{r, class.source="bg-danger"}

```

```{r, class.source="bg-success"}
rows_with_missing_values <- df[!complete.cases(df), ]
```

Don't use number suffixes to distinguish variables

## Use one word for one concept

`get`, `retrieve`, `fetch` are synonyms. Pick one if two functions perform
the same action.

```{r, class.source="bg-danger"}
get_client_data <- function(...)
fetch_location_data <- function(...)
```

```{r, class.source="bg-success"}
get_client_data <- function(...)
get_location_data <- function(...)
```

## Use verbs to name functions

Functions _do_ something with inputs. As such, choose a name that reflects what it
_does_.

```{r, class.source="bg-danger"}
client_data <- function(...)
```

```{r, class.source="bg-success"}
get_client_data <- function(...)
```

## Do not overwrite variables

```{r, class.source="bg-danger"}
customers <- get_rid_of_rows_with_missing_values(customers)
```

```{r, class.source="bg-success"}
complete_customers <- get_rid_of_rows_with_missing_values(customers)
```

## Choose names that do not conflict with base functions or keywords

# 2. Functions

## Don't repeat yourself (DRY)

## Extract code to function to name it

Even if you don't plan to reuse the code in more than one places.

## Avoid too many parameters (> 3)

## Pass all parameters to a function as arguments.

This makes functions self-contained. Very rare exception: global constants
with a naming convention that is easy to follow (e.g. `ALL_CAPITAL_LETTERS`).

## Clearly separate functions with side effects and functions with a return value

## Organize your functions from top to down in abstraction levels.

Main functions should come first, lower level functions that they use come
below them.

# 3. Comments

## Explain with the code itself rather than with comments

```{r, class.source="bg-danger"}
# calculate customer lifetime value
c_ltv <- calc_cust_LTV(cust_data)
```

```{r, class.source="bg-success"}
customer_lifetime_value <- calculate_lifetime_value(customer)
```

# 4. General refactoring tips

## First write a working code, then make it cleaner (= refactor)

## Boy scout rule: if you modify something, think about leaving it cleaner than it was

If you have to touch a piece of code for any reason, consider refactoring it as well.

## Code should be readable also to people not familiar with R

You don't have to overdo refactoring (dont' create a new function for `dplyr::filter`),
however, hide somewhat cryptic parts to named piece of code (example: `apply(dt, 2, fun)`)

# Literature

Robert C. Martin: Clean code.
